ms_0_intro

### 2018-03-13 man/ms_0_intro.Rmd
update intro

# 1) Sequías y Land use change

Drought is a natural phenomenon occurs when water availability is significantly below-normal levels [@Dai2011]. It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Dai2011; @Ummenhofer2017]. Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forest through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017].

In the last decades there is an crescent interest in assessing the drought impacts on ecosystem, especially considering some studies that suggest an increase in severity, geographic extent and frequency of droughts as a result of climate change [@Dai2011; @Trenberth2014]. There are studies analyzing the response of vegetation to drought from global (*e.g*. @Allen2010; @VicenteSerrano2013; @Norman2016) to local scales (*e.g* @MartinezVilalta2016 for a revision). Some studies have focusing on the drought effects on Mediterranean species which representing their southernmost limit [@Pasho2011; @Camarero2011; @Herrero2014; @Linares2014; @SanchezSalguero2017; @Caminero2018].

Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b]. In addition, climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013] particularly for the Mediterranean area [@Spinoni2017a]. In the Iberian Peninsula, major drought episodies were recorded in 1943, 1981, 1995, 2000 and 2005 [@VicenteSerrano2014; @Guerreiro2017]. The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017]. These events were extreme in both its magnitude and spatial extent [@Gouveia2014].

Vegetation responses to drought are influenced by other drivers of environmental change [@Fischer2006; @Oliver2014; @Franklin2016; @Penuelas2017]. The impacts of drought on ecosystem must be evaluated in a context of global change considering the interactions with other drivers of environmental change [@Archaux2006; @Clavero2011], such us land-use change [@DoblasMiranda2017]. This is specially relevant for areas with a long history of landscape modification such us Mediterranean region where land-use change is a key driver of the global change [@Navarro2013; @Ameztegui2016]. In a recent review, @DoblasMiranda2017 found that land use changes and more especially drought, are crucial factors in the interactions among the drivers of global change for Mediterranean forests.

The impacts of drought are especially relevant for populations located in the retreating or rear edge of their distribution where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010].

There are several works que han analizado la respuesta de la vegetación a la sequía en su rear edge, y en este contexto es interesante analizar no solo el efecto de la sequía en estas poblaciones sino también su resiliencia. -- Resilience is the capacity of an ecosystem, community or individual to restore pre-disturbance structure and function after a disturbance (analogous to ‘engineering resilience’, see Holling, 1996; see also Lloret et al., 2011). Thus, the assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services. Furthermore, the analysis of resilience using multiple demographic rates allows us to detect key vulnerabilities to drought in tree population dynamics, such as reductions in recruitment or survival, which can prompt a decline in dominant species shaping forest structure.

@@

@@

b) Remote sensing:

Remote sensing can be used for studying droughts impacts on ecosystems (*e.g.* @Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016). Several studies have analyzed drought impacts on vegetation using remote sensing at different scales [@VicenteSerrano2013; @Panisset2018], and for Mediterranean area there were studies focused on local scales (*e.g* @Lloret2007; @Gouveia2012, @Zribi2016) and recently other covering all Mediterranean region [@Gouveia2017].

sin embargo (aqui poner las limitaciones de una u otra metodología).

Existen algunos estudios que han combinado el uso de dendro y RS para estudiar el efecto de las sequías....

* @Dorman2013 --> interesante compara NDVI y dendro 
* Tree growth and vegetation activity at the ecosystem-scale in the eastern Mediterranean

Aqui nosotros evaluamos el efecto de la sequía ... The vegetation response to water deficit was assessed using remote sensing information and dendrocronologial data.

... ojo los remote sensing datos (aun con alta resolucion temporal y/o espacial) son insuficientes por si solos para demostrar 
 Remote sensing data, even high temporal or spatial resolution data, are insufficient alone to unequivocally demonstrate that drought is the causal agent of a particular change in reflectance values

@@

### ojo un parrafito de resiliencia

Shifts in forest structure and composition in response to drought could alter ecosystem functions profoundly, affecting hydrological processes (Allen, 2007; Adams et al., 2012), nutrient cycling (Hughes et al., 2006; Xiong et al., 2011) and land-surface properties (Royer et al., 2011; Anderegg et al., 2013).

In this context, resilience is an important concept that is gaining unprecedented attention in ecology (e.g. Reyer et al., 2015). Resilience is the capacity of an ecosystem, community or individual to restore pre-disturbance structure and function after a disturbance (analogous to ‘engineering resilience’, see Holling, 1996; see also Lloret et al., 2011). Thus, the assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services. Furthermore, the analysis of resilience using multiple demographic rates allows us to detect key vulnerabilities to drought in tree population dynamics, such as reductions in recruitment or survival, which can prompt a decline in dominant species shaping forest structure.

@@

>>>>

Drought is a natural phenomenon occurring when water availability is significantly below-normal levels [@Dai2011]. It is one of the most frequent and widespread climatic extremes affecting ecological systems across every climatic zone worldwide [@Mishra2010; @Dai2011; @Ummenhofer2017]. Summer drought is a characteristic feature of the Mediterranean climate [@Lionello2012], but an increase in the frequency and severity of drought events have been recorded for the Mediterranean region [@Hoerling2011], particularly for southern Europe [@VicenteSerrano2014; @Spinoni2015; @Stagge2017] where a trend towards drier summers has been reported for the last three decades [@Spinoni2017b]. Climate change projections suggest that extreme events, like droughts, will become more frequent and severe [@Trenberth2014; @IPCC2013]  @Spinoni2017a], specially in the Mediterranean area [@Spinoni2017a].

Drought causes significant changes in ecosystem productivity and water dynamics that can severely affect forests through changes in plant physiology, phenology and carbon allocation [@Allen2010; @Ummenhofer2017]. In the last decades the assessment of the drought impacts on ecosystems has gained attention [@Allen2010; @Clark2016], especially considering some studies that suggest an increase in severity, geographic extent and frequency of droughts as a result of climate change [@Dai2011; @Trenberth2014]. The response of vegetation to drought has been analyzed at different scales, from global [*e.g*. @Allen2010; @VicenteSerrano2013; @Norman2016] to local scales [see @MartinezVilalta2016 for a revision].

Vegetation responses to drought are influenced by other drivers of environmental change [@Fischer2006; @Oliver2014; @Franklin2016; @Penuelas2017], and some authors states that the impacts of drought on ecosystem must be evaluated in a context of global change considering the interactions with other drivers of environmental change [@Archaux2006; @Clavero2011], such us land-use change [@DoblasMiranda2017]. This is specially relevant for areas with a long history of landscape modification as the Mediterranean region where land-use change is a key driver of the global change [@Navarro2013; @Ameztegui2016]. For instance, @DoblasMiranda2017 in a recent review, found that land use changes and more especially drought, are crucial factors in the interactions among the drivers of global change for Mediterranean forests.

The impacts of drought are also especially relevant for populations located in the rear edge of their distribution, where climatic conditions often corresponds to the species-limits and the populations are likey to be more sensitive to small variations on climate [@Hampe2005]. The response of rear-edge populations to variations on climate could be useful to estimate the adaptation of the species to predicted climatic changes [@Jump2010]. It becomes even more important for relict populations driven by historical land-uses, which can harbour high levels of intraspecific genetic diversity [@Valbuena2010].

Several studies analyzed the drought effects on Mediterranean tree species representing the southermost limit of their distribution [*e.g*. @Pasho2011; @Camarero2011; @SanchezSalguero2012; @Linares2014; @Dorado2017b; @SanchezSalguero2017; @Caminero2018; ...], however only a few works have focused on the resilience to drought of rear-edge populations [*e.g.* @SanchezSalguero2013; @Herrero2014]. Resilience is the capacity of an ecosystem, community or individual to recover pre-disturbance structure and function after a disturbance (@Holling1973; @Lloret2011; @Hodgson2015). The assessment of forest resilience to climatic disturbances, such as extreme droughts, provides critical information about the capacity of forests to maintain their structure and to continue providing valuable ecosystem services. This is particularly relevant for populations located in their rear-edge of their distribution, especially when they are located in Mediterranean mountains.

In this work we combined remote sensing information and dendroecological methods to evaluate the drought impacts in canopy greenness (as a proxy to primary growth) and radial growth of a Mediterranean tree species (*Quercus pyrenaica*) located in the rear-edge of their distribution. We also assessed the resilience of these forest to severe drought events.

Sabemos que la sequía afecta al crecimiento y a la productividad (poner ejemplos de ambos trabajos) ..

RS 
Remote sensing can be used for studying droughts impacts on ecosystems [*e.g.* @Deshayes2006; @Zhang2013; @AghaKouchak2015; @McDowell2015; @Norman2016]. Drought impacts on vegetation using remote sensing have been analyzed at different scales [@VicenteSerrano2013; @Panisset2018; @Assal2016]. For Mediterranean area there were studies covering all region [@Gouveia2017] and others focused on local scales [*e.g* @Lloret2007; @Gouveia2012, @Zribi2016].

hablar algo de las ventajas de la RS ...

Existen algunos estudios que han combinado el uso de dendro y RS para estudiar el efecto de las sequías.... 
* @Dorman2013 --> interesante compara NDVI y dendro 
* Tree growth and vegetation activity at the ecosystem-scale in the eastern Mediterranean

e incluso otros trabajos han analizado RS y dendro para evaluar resiliencia, pero han utilizado datos de amplia escala (Gazol et al 2018) y no se han centrado en rear edge.

Por otro lado

Dendro

@@

In the Iberian Peninsula, major drought episodies were recorded in 1943, 1981, 1995, 2000 and 2005 [@VicenteSerrano2014; @Guerreiro2017]. The 2004/2005 and 2011/2012 hydrological years are considered two of the worst drought periods recorded in the Iberian Peninsula, particularly in the southern sector [@GarciaHerrera2007; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017]. These events were extreme in both its magnitude and spatial extent [@Gouveia2014].

@@

@@

@@

--> plantear la hipótesis... efecto de sequias recurrentes ...

y luego los objetivos
